---
title: "Exploring Denver ACS Data"
author: "Molly Graber"
date: "11/19/2017"
output: html_document
---
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(leaflet)
library(leaflet.extras)
library(rgdal)
library(rgeos)
```

##Introduction

The following maps present different methods of visualizing American Community Surver data of Denver using leaflet, both from a variable perspective and a contextual perspective. The data come from 2009-2013 5-year estimates at the tract level, availible through the Denver Open Data catalog. 

```{r, echo=FALSE, warning=FALSE, results='hide'}
#tracts <- readOGR('ACS1014_tract', layer = 'ACS1014_tract')
#ahu <- readOGR('affordable_housing_units', layer = 'affordable_housing_units')

tracts<- readOGR('ACS', layer = 'ACS')
race <- c('PCT_HISPAN','PCT_WHITE','PCT_BLACK','PCT_NATIVE','PCT_ASIAN','PCT_HAWAII','PCT_OTHERR')
age <- c('AGE_0_TO_9','AGE_10_TO1','AGE_20_TO_','AGE_30_TO_','AGE_40_TO_','AGE_50_TO_','AGE_60_TO_','AGE_70_TO_','AGE_80_PLU')
edu <- c('HSGRAD_OR_EQUIV_EDU','BACHELORS_OR_HIGHER_EDU')
lang <- c('ONLY_ENGLISH_LNG','SPANISH_LNG')
family <- c('MARRIED_CO','FAMILY_HOU','NONFAMILY_')
wealth <- c('MED_HH_INC')
columns <- c(race, age, edu, lang, family, wealth)
```

##Univariate Maps

Map of Median Household Income by Census Tract
```{r interactive choropleth, echo=FALSE}
pal <- colorBin("BuGn", domain = tracts$MED_HH_INC, n=6)

labels <- sprintf(
  "<strong>Median Household Income</strong><br/>$%g ",
  tracts$MED_HH_INC
) %>% lapply(htmltools::HTML)

leaflet(tracts) %>%
  setView(-104.99, 39.73, zoom = 11) %>%
  addProviderTiles(providers$CartoDB.Positron, group="Light") %>%
  addPolygons(
    fillColor = ~pal(tracts$MED_HH_INC),
    weight = 2,
    opacity = .3,
    color = "white",
    dashArray = "",
    fillOpacity = 0.4,
    highlight = highlightOptions(
      weight = 3,
      color = "#666",
      dashArray = "",
      fillOpacity = 1,
      bringToFront = TRUE),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) %>%
  addLegend(pal = pal, values = tracts$MED_HH_INC, opacity = 1, title = "Median Household Income ($)",
    position = "bottomleft")

```

Map of Ethnic Populations by Census Tract
```{r hispanic choropleth, echo=FALSE}

palHispanic <- colorBin("Purples", domain = tracts$PCT_HISPAN, 5)
palBlack <- colorBin("Blues", domain = tracts$PCT_BLACK, 5)
palAsian <- colorBin("Greens", domain = tracts$PCT_ASIAN, 5)

labelsH <- sprintf(
  "<strong>Percent Hispanic</strong><br/>%g Percent ",
  tracts$PCT_HISPAN
) %>% lapply(htmltools::HTML)

labelsB <- sprintf(
  "<strong>Percent Black</strong><br/>%g Percent ",
  tracts$PCT_BLACK
) %>% lapply(htmltools::HTML)

labelsA <- sprintf(
  "<strong>Percent Asian</strong><br/>%g Percent ",
  tracts$PCT_ASIAN
) %>% lapply(htmltools::HTML)


leaflet(tracts) %>%
  setView(-104.99, 39.73, zoom = 11) %>%
  addProviderTiles(providers$CartoDB.Positron, group="Light") %>%
  addPolygons(
    fillColor = ~palHispanic(tracts$PCT_HISPAN),
    group = "Hispanic",
    weight = 2,
    opacity = .3,
    color = "white",
    dashArray = "",
    fillOpacity = 0.6,
    highlight = highlightOptions(
      weight = 3,
      color = "#666",
      dashArray = "",
      fillOpacity = 1,
      bringToFront = TRUE),
    label = labelsH,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) %>%
  addPolygons(
    fillColor = ~palBlack(tracts$PCT_BLACK),
    group = "Black",
    weight = 2,
    opacity = .3,
    color = "white",
    dashArray = "",
    fillOpacity = 0.6,
    highlight = highlightOptions(
      weight = 3,
      color = "#666",
      dashArray = "",
      fillOpacity = 1,
      bringToFront = TRUE),
    label = labelsB,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) %>%
  addPolygons(
    fillColor = ~palAsian(tracts$PCT_ASIAN),
    group = "Asian",
    weight = 2,
    opacity = .3,
    color = "white",
    dashArray = "",
    fillOpacity = 0.6,
    highlight = highlightOptions(
      weight = 3,
      color = "#666",
      dashArray = "",
      fillOpacity = 1,
      bringToFront = TRUE),
    label = labelsA,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) %>%
  addLayersControl(
    baseGroups = c("Hispanic","Black","Asian"),
    options = layersControlOptions(collapsed = FALSE)
  )



```


Z-score Standardized Map of Median Household Income

```{r standardized choropleth, echo=FALSE}

palZ <- colorBin("BrBG", domain = scale(tracts$MED_HH_INC), c(-10,-3, -1, 0, 1, 3, 10))


labels <- sprintf(
  "<strong>Median Household Income Z-score</strong><br/>%g SEs from Denver Mean ",
  scale(tracts$MED_HH_INC)
) %>% lapply(htmltools::HTML)

leaflet(tracts) %>%
  setView(-104.99, 39.73, zoom = 11) %>%
  addProviderTiles(providers$CartoDB.Positron, group="Light") %>%
  addPolygons(
    fillColor = ~palZ(scale(tracts$MED_HH_INC)),
    weight = 2,
    opacity = .3,
    color = "white",
    dashArray = "",
    fillOpacity = 0.4,
    highlight = highlightOptions(
      weight = 3,
      color = "#666",
      dashArray = "",
      fillOpacity = 1,
      bringToFront = TRUE),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) %>%
  addLegend(pal = palZ, values = scale(tracts$MED_HH_INC), opacity = 1, title = "Med. Income Z-score",
    position = "bottomleft")


```

Map of K-Means for 3, 4, and 5 Clusters

```{r, echo=FALSE}
cluster3 <- kmeans(scale(tracts@data[,8:100]), 3, iter.max=10000, nstart=100)
tracts@data$cluster3 <- as.factor(cluster3$cluster)

cluster4 <- kmeans(scale(tracts@data[,8:100]), 4, iter.max=10000, nstart=100)
tracts@data$cluster4 <- as.factor(cluster4$cluster)

cluster5 <- kmeans(scale(tracts@data[,8:100]), 5, iter.max=10000, nstart=100)
tracts@data$cluster5 <- as.factor(cluster5$cluster)

cluster6 <- kmeans(scale(tracts@data[,8:100]), 6, iter.max=10000, nstart=100)
tracts@data$cluster6 <- as.factor(cluster6$cluster)

cluster7 <- kmeans(scale(tracts@data[,8:100]), 7, iter.max=10000, nstart=100)
tracts@data$cluster7 <- as.factor(cluster7$cluster)

cluster8 <- kmeans(scale(tracts@data[,8:100]), 8, iter.max=10000, nstart=100)
tracts@data$cluster8 <- as.factor(cluster8$cluster)

cluster9 <- kmeans(scale(tracts@data[,8:100]), 9, iter.max=10000, nstart=100)
tracts@data$cluster9 <- as.factor(cluster9$cluster)

cluster10 <- kmeans(scale(tracts@data[,8:100]), 10, iter.max=10000, nstart=100)
tracts@data$cluster10 <- as.factor(cluster10$cluster)

```

```{r, echo=FALSE}

leaf <- leaflet(tracts) %>%
  addProviderTiles(providers$CartoDB.Positron)

### Set up 6-cluster layer
pal3 <- colorFactor("Set3", domain = tracts$cluster3)
pal4 <- colorFactor("Set3", domain = tracts$cluster4)
pal5 <- colorFactor("Set3", domain = tracts$cluster5)


labels3 <- sprintf(
  "<strong>Tract Demographic Type</strong><br/>%g",
  tracts$cluster3
) %>% lapply(htmltools::HTML)

labels4 <- sprintf(
  "<strong>Tract Demographic Type</strong><br/>%g",
  tracts$cluster4
) %>% lapply(htmltools::HTML)

labels5 <- sprintf(
  "<strong>Tract Demographic Type</strong><br/>%g",
  tracts$cluster5
) %>% lapply(htmltools::HTML)


## Add polygon groups
leaf %>%
  addPolygons(
      group = "3 Clusters",
      fillColor = ~pal3(tracts$cluster3),
      weight = 2,
      opacity = .3,
      color = "white",
      dashArray = "",
      fillOpacity = 0.4,
      highlight = highlightOptions(
        weight = 3,
        color = "#666",
        dashArray = "",
        fillOpacity = 1,
        bringToFront = TRUE),
      label = labels3,
      labelOptions = labelOptions(
        style = list("font-weight" = "normal", padding = "3px 8px"),
        textsize = "15px",
        direction = "auto")) %>%
  addPolygons(
      group = "4 Clusters",
      fillColor = ~pal4(tracts$cluster4),
      weight = 2,
      opacity = .3,
      color = "white",
      dashArray = "",
      fillOpacity = 0.4,
      highlight = highlightOptions(
        weight = 3,
        color = "#666",
        dashArray = "",
        fillOpacity = 1,
        bringToFront = TRUE),
      label = labels4,
      labelOptions = labelOptions(
        style = list("font-weight" = "normal", padding = "3px 8px"),
        textsize = "15px",
        direction = "auto")) %>%
  addPolygons(
      group = "5 Clusters",
      fillColor = ~pal5(tracts$cluster5),
      weight = 2,
      opacity = .3,
      color = "white",
      dashArray = "",
      fillOpacity = 0.4,
      highlight = highlightOptions(
        weight = 3,
        color = "#666",
        dashArray = "",
        fillOpacity = 1,
        bringToFront = TRUE),
      label = labels5,
      labelOptions = labelOptions(
        style = list("font-weight" = "normal", padding = "3px 8px"),
        textsize = "15px",
        direction = "auto")) %>%
  setView(-104.99, 39.73, zoom = 12) %>%
  addLayersControl(
    baseGroups = c("3 Clusters","4 Clusters","5 Clusters"),
    options = layersControlOptions(collapsed = FALSE)
  )

```

